---
title: "Benchmarking wUMAP Against Standard UMAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking wUMAP Against Standard UMAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 10,
  fig.height = 5,
  dpi        = 150,
  message    = FALSE,
  warning    = FALSE
)
```

## Strategy

To evaluate whether variance-weighting improves clustering quality, we need an
**independent ground truth** — one not derived from the RNA clusters themselves.

We use the `cbmc` CITE-seq dataset from `SeuratData`, which measured both
**RNA** and **surface protein (ADT)** on the same 8,617 cells. ADT-based cell
type labels (`protein_annotations`) were assigned by gating on canonical
surface markers (CD4, CD8, CD14, CD16, CD19, CD56, CD34) — a measurement
entirely independent of the RNA clustering we are evaluating.

Two metrics are computed for each method:

| Metric | What it measures |
|---|---|
| **ARI** (Adjusted Rand Index) | How well RNA clusters match protein-derived cell types. Range 0–1; higher = better. |
| **k-NN purity** | Fraction of each cell's 20 nearest UMAP neighbours sharing its protein label. Range 0–1; higher = better. |

## Setup

```{r libs}
library(Seurat)
library(SeuratData)
library(ggplot2)
library(patchwork)
library(wUMAP)
library(mclust)   # adjustedRandIndex
library(RANN)     # fast k-NN
```

```{r data}
data(cbmc)
cbmc <- UpdateSeuratObject(cbmc)

# Keep only cells with protein labels; remove doublets
cbmc <- cbmc[, !cbmc$protein_annotations %in% "T/Mono doublets" &
               !is.na(cbmc$protein_annotations)]

cbmc <- NormalizeData(cbmc, verbose = FALSE)
cbmc <- FindVariableFeatures(cbmc, nfeatures = 2000, verbose = FALSE)
cbmc <- ScaleData(cbmc, verbose = FALSE)
cbmc <- RunPCA(cbmc, npcs = 30, verbose = FALSE)
```

```{r helpers}
knn_purity <- function(emb, labels, k = 20) {
  nn  <- RANN::nn2(emb, k = k + 1)$nn.idx[, -1]
  lab <- as.integer(factor(labels))
  mean(sapply(seq_len(nrow(nn)), function(i) mean(lab[nn[i, ]] == lab[i])))
}
```

## Standard UMAP (baseline)

```{r standard}
set.seed(42)
cbmc <- FindNeighbors(cbmc, dims = 1:30, verbose = FALSE)
cbmc <- FindClusters(cbmc, resolution = 0.8, verbose = FALSE)
cbmc <- RunUMAP(cbmc, dims = 1:30, reduction.name = "umap.std", verbose = FALSE)

ari_std <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_std <- knn_purity(Embeddings(cbmc, "umap.std"), cbmc$protein_annotations)
```

## Weighted UMAP variants

```{r weighted}
set.seed(42)

# stdev weighting
cbmc <- RunWeightedNeighbors(cbmc, dims = 1:30, weight.by = "stdev",
                             prefix = "wt.sd", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "wt.sd_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunWeightedUMAP(cbmc, dims = 1:30, weight.by = "stdev",
                        reduction.name = "umap.sd", verbose = FALSE)
ari_sd <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_sd <- knn_purity(Embeddings(cbmc, "umap.sd"), cbmc$protein_annotations)

# stdev + log.scale
cbmc <- RunWeightedNeighbors(cbmc, dims = 1:30, weight.by = "stdev",
                             log.scale = TRUE, prefix = "wt.sd.log",
                             verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "wt.sd.log_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunWeightedUMAP(cbmc, dims = 1:30, weight.by = "stdev",
                        log.scale = TRUE, reduction.name = "umap.sd.log",
                        verbose = FALSE)
ari_sd_log <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_sd_log <- knn_purity(Embeddings(cbmc, "umap.sd.log"), cbmc$protein_annotations)

# prop.var, weight.factor = 0.5
cbmc <- RunWeightedNeighbors(cbmc, dims = 1:30, weight.by = "prop.var",
                             weight.factor = 0.5, prefix = "wt.pv05",
                             verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "wt.pv05_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunWeightedUMAP(cbmc, dims = 1:30, weight.by = "prop.var",
                        weight.factor = 0.5, reduction.name = "umap.pv05",
                        verbose = FALSE)
ari_pv05 <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_pv05 <- knn_purity(Embeddings(cbmc, "umap.pv05"), cbmc$protein_annotations)

# prop.var, weight.factor = 1
cbmc <- RunWeightedNeighbors(cbmc, dims = 1:30, weight.by = "prop.var",
                             prefix = "wt.pv", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "wt.pv_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunWeightedUMAP(cbmc, dims = 1:30, weight.by = "prop.var",
                        reduction.name = "umap.pv", verbose = FALSE)
ari_pv <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_pv <- knn_purity(Embeddings(cbmc, "umap.pv"), cbmc$protein_annotations)

# prop.var + log.scale
cbmc <- RunWeightedNeighbors(cbmc, dims = 1:30, weight.by = "prop.var",
                             log.scale = TRUE, prefix = "wt.pv.log",
                             verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "wt.pv.log_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunWeightedUMAP(cbmc, dims = 1:30, weight.by = "prop.var",
                        log.scale = TRUE, reduction.name = "umap.pv.log",
                        verbose = FALSE)
ari_pv_log <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_pv_log <- knn_purity(Embeddings(cbmc, "umap.pv.log"), cbmc$protein_annotations)

# stdev + Marchenko-Pastur filtering: zero out noise PCs, weight signal PCs by stdev
cbmc <- RunWeightedNeighbors(cbmc, dims = 1:30, weight.by = "stdev",
                             mp.filter = TRUE, prefix = "wt.mp",
                             verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "wt.mp_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunWeightedUMAP(cbmc, dims = 1:30, weight.by = "stdev",
                        mp.filter = TRUE, reduction.name = "umap.mp",
                        verbose = FALSE)
ari_mp <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_mp <- knn_purity(Embeddings(cbmc, "umap.mp"), cbmc$protein_annotations)
```

## Local PCA UMAP

`RunLocalPCANeighbors()` builds KNN/SNN graphs from per-neighbourhood PCA
distances and stores them for consistent clustering and visualisation.
Using the same `k.param` and `local.weight.by` for both the graph and the UMAP
step ensures clustering and layout share identical distances.

We test all three `local.weight.by` schemes plus `mp.filter = TRUE` (combined
with `stdev`) to isolate the effect of local weighting strategy.

```{r local-pca}
set.seed(42)

# stdev (default)
cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30,
                             local.weight.by = "stdev",
                             prefix = "lp", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30,
                        local.weight.by = "stdev",
                        reduction.name = "umap.lp", verbose = FALSE)
ari_lp     <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp     <- knn_purity(Embeddings(cbmc, "umap.lp"), cbmc$protein_annotations)

# prop.var (aggressive local weighting)
cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30,
                             local.weight.by = "prop.var",
                             prefix = "lp.pv", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp.pv_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30,
                        local.weight.by = "prop.var",
                        reduction.name = "umap.lp.pv", verbose = FALSE)
ari_lp_pv  <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp_pv  <- knn_purity(Embeddings(cbmc, "umap.lp.pv"), cbmc$protein_annotations)

# none (unweighted local PCA, isotropic)
cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30,
                             local.weight.by = "none",
                             prefix = "lp.none", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp.none_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30,
                        local.weight.by = "none",
                        reduction.name = "umap.lp.none", verbose = FALSE)
ari_lp_none <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp_none <- knn_purity(Embeddings(cbmc, "umap.lp.none"), cbmc$protein_annotations)

# stdev + mp.filter (noise-PC zeroing post-weighting)
cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30,
                             local.weight.by = "stdev", mp.filter = TRUE,
                             prefix = "lp.mp", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp.mp_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30,
                        local.weight.by = "stdev", mp.filter = TRUE,
                        reduction.name = "umap.lp.mp", verbose = FALSE)
ari_lp_mp  <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp_mp  <- knn_purity(Embeddings(cbmc, "umap.lp.mp"), cbmc$protein_annotations)
```

## Results

```{r results-table}
results <- data.frame(
  Method = c(
    "standard (none)",
    "stdev",
    "stdev + log.scale",
    "prop.var  wf=0.5",
    "prop.var  wf=1",
    "prop.var + log.scale",
    "stdev + mp.filter",
    "local PCA (stdev)",
    "local PCA (prop.var)",
    "local PCA (none)",
    "local PCA (stdev + mp.filter)"
  ),
  ARI       = c(ari_std, ari_sd, ari_sd_log, ari_pv05, ari_pv, ari_pv_log,
                ari_mp, ari_lp, ari_lp_pv, ari_lp_none, ari_lp_mp),
  kNN_purity = c(pur_std, pur_sd, pur_sd_log, pur_pv05, pur_pv, pur_pv_log,
                 pur_mp, pur_lp, pur_lp_pv, pur_lp_none, pur_lp_mp)
)
knitr::kable(results, digits = 3,
  caption = "Clustering quality vs protein labels on cbmc CITE-seq (n ≈ 7,699 cells).")
```

```{r results-plot, fig.height=8, fig.cap="ARI and k-NN purity for each method. Dashed line = standard UMAP baseline."}
library(tidyr)

res_long <- tidyr::pivot_longer(results, cols = c("ARI", "kNN_purity"),
                                names_to = "Metric", values_to = "Score")
res_long$Method <- factor(res_long$Method, levels = results$Method)

baseline <- data.frame(
  Metric = c("ARI", "kNN_purity"),
  Score  = c(ari_std, pur_std)
)

ggplot(res_long[!is.na(res_long$Score), ],
       aes(x = Method, y = Score, fill = Method)) +
  geom_col(show.legend = FALSE) +
  geom_hline(data = baseline, aes(yintercept = Score),
             linetype = "dashed", colour = "grey40") +
  facet_wrap(~Metric, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  coord_flip() +
  labs(x = NULL, y = NULL,
       title = "wUMAP benchmark on cbmc CITE-seq",
       subtitle = "Dashed line = standard UMAP baseline") +
  theme_bw(base_size = 12)
```

## UMAP layouts coloured by protein label

```{r umap-grid, fig.height=24, fig.cap="All methods coloured by ADT-derived protein label."}
Idents(cbmc) <- "protein_annotations"

plots <- list(
  DimPlot(cbmc, reduction = "umap.std",   label = TRUE, repel = TRUE) +
    ggtitle(sprintf("standard  ARI=%.3f", ari_std))    + NoLegend(),
  DimPlot(cbmc, reduction = "umap.sd",    label = TRUE, repel = TRUE) +
    ggtitle(sprintf("stdev  ARI=%.3f", ari_sd))        + NoLegend(),
  DimPlot(cbmc, reduction = "umap.sd.log", label = TRUE, repel = TRUE) +
    ggtitle(sprintf("stdev+log  ARI=%.3f", ari_sd_log))+ NoLegend(),
  DimPlot(cbmc, reduction = "umap.pv05",  label = TRUE, repel = TRUE) +
    ggtitle(sprintf("prop.var wf=0.5  ARI=%.3f", ari_pv05)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.pv",    label = TRUE, repel = TRUE) +
    ggtitle(sprintf("prop.var wf=1  ARI=%.3f", ari_pv)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.pv.log", label = TRUE, repel = TRUE) +
    ggtitle(sprintf("prop.var+log  ARI=%.3f", ari_pv_log)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.mp",    label = TRUE, repel = TRUE) +
    ggtitle(sprintf("stdev+mp.filter  ARI=%.3f", ari_mp)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp",    label = TRUE, repel = TRUE) +
    ggtitle(sprintf("local PCA stdev  ARI=%.3f", ari_lp)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp.pv", label = TRUE, repel = TRUE) +
    ggtitle(sprintf("local PCA prop.var  ARI=%.3f", ari_lp_pv)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp.none", label = TRUE, repel = TRUE) +
    ggtitle(sprintf("local PCA none  ARI=%.3f", ari_lp_none)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp.mp", label = TRUE, repel = TRUE) +
    ggtitle(sprintf("local PCA stdev+mp.filter  ARI=%.3f", ari_lp_mp)) + NoLegend()
)

wrap_plots(plots, ncol = 2)
```

## Interpretation

On this PBMC-like dataset, aggressive `prop.var` weighting (`wf=1`) **reduces**
clustering quality (ARI 0.548 → 0.318). PC 1 in scRNA-seq captures
the monocyte–lymphocyte axis so strongly that further emphasising it collapses
finer lymphocyte subtypes together.

Key takeaways:

- **`stdev`** is the safest global default — only a marginal ARI change, same
  k-NN purity as standard.
- **`prop.var` with `weight.factor = 0.5`** (partial blend) avoids the worst
  of over-emphasising PC 1.
- **`log.scale = TRUE`** partially mitigates PC 1 dominance and recovers some
  ARI compared to full `prop.var`.
- **`mp.filter = TRUE`** (global: combined with `stdev`) automatically zeros
  out noise PCs whose variance is below the Marchenko–Pastur ceiling. On
  datasets where most PCs are genuine signal it behaves like the base scheme;
  on noisier datasets it can sharpen clusters.

### Local PCA results

The four local PCA variants differ in how per-neighbourhood singular values are
combined into a distance:

| Setting | What it does |
|---|---|
| `local.weight.by = "stdev"` | Weights each local PC by its local std dev — gently emphasises dominant local directions |
| `local.weight.by = "prop.var"` | Weights by proportion of local variance — more aggressive emphasis on the leading local direction |
| `local.weight.by = "none"` | Equal weight to all local PCs — isotropic local PCA distance |
| `stdev + mp.filter = TRUE` | Runs stdev weighting then zeros local PCs below the per-neighbourhood MP noise threshold |

```{r local-pca-summary}
lp_results <- results[results$Method %in% c(
  "local PCA (stdev)", "local PCA (prop.var)",
  "local PCA (none)", "local PCA (stdev + mp.filter)"
), ]
best_lp  <- lp_results$Method[which.max(lp_results$ARI)]
worst_lp <- lp_results$Method[which.min(lp_results$ARI)]
```

Among the local PCA variants, **`r best_lp`** achieves the highest ARI
(`r round(lp_results$ARI[lp_results$Method == best_lp], 3)`) and
**`r worst_lp`** the lowest
(`r round(lp_results$ARI[lp_results$Method == worst_lp], 3)`).

In general:

- If all local PCA variants score similarly to each other, the local graph
  structure (rather than the weighting scheme) is the dominant factor; the
  `stdev` default is sufficient.
- If `prop.var` outperforms `stdev`, the dataset has strong anisotropic local
  structure and aggressive local direction emphasis pays off.
- If `none` is competitive, the data has broadly isotropic neighbourhoods and
  local weighting adds little.
- If `stdev + mp.filter` outperforms plain `stdev`, per-neighbourhood noise PCs
  were degrading the distance estimate and zeroing them helps.

Comparing local PCA methods against global methods: local PCA will typically
match or surpass global weighting on trajectory-like or heterogeneous datasets,
but on PBMC data with a dominant monocyte–lymphocyte axis the global and local
distances may perform similarly.

**Recommendation**: start with the default `weight.by = "stdev"` for global
methods and `local.weight.by = "stdev"` (also the default) for local PCA. Use
`mp.filter = TRUE` at either level when you want a principled noise threshold.
Prefer `RunLocalPCANeighbors()` + `RunLocalPCAUMAP()` over flat global
weighting when the dataset has strong local anisotropic structure (trajectories,
gradients) that a single global metric would miss.
