---
title: "Benchmarking wUMAP Against Standard UMAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking wUMAP Against Standard UMAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 10,
  fig.height = 5,
  dpi        = 150,
  message    = FALSE,
  warning    = FALSE
)
```

## Strategy

To evaluate whether variance-weighting improves clustering quality, we need an
**independent ground truth** — one not derived from the RNA clusters themselves.

We use the `cbmc` CITE-seq dataset from `SeuratData`, which measured both
**RNA** and **surface protein (ADT)** on the same 8,617 cells. ADT-based cell
type labels (`protein_annotations`) were assigned by gating on canonical
surface markers (CD4, CD8, CD14, CD16, CD19, CD56, CD34) — a measurement
entirely independent of the RNA clustering we are evaluating.

Two metrics are computed for each method:

| Metric | What it measures |
|---|---|
| **ARI** (Adjusted Rand Index) | How well RNA clusters match protein-derived cell types. Range 0–1; higher = better. |
| **k-NN purity** | Fraction of each cell's 20 nearest UMAP neighbours sharing its protein label. Range 0–1; higher = better. |

## Setup

```{r libs}
library(Seurat)
library(SeuratData)
library(ggplot2)
library(patchwork)
library(wUMAP)
library(mclust)   # adjustedRandIndex
library(RANN)     # fast k-NN
```

```{r data}
data(cbmc)
cbmc <- UpdateSeuratObject(cbmc)

# Keep only cells with protein labels; remove doublets
cbmc <- cbmc[, !cbmc$protein_annotations %in% "T/Mono doublets" &
               !is.na(cbmc$protein_annotations)]

cbmc <- NormalizeData(cbmc, verbose = FALSE)
cbmc <- FindVariableFeatures(cbmc, nfeatures = 2000, verbose = FALSE)
cbmc <- ScaleData(cbmc, verbose = FALSE)
cbmc <- RunPCA(cbmc, npcs = 30, verbose = FALSE)
```

```{r helpers}
knn_purity <- function(emb, labels, k = 20) {
  nn  <- RANN::nn2(emb, k = k + 1)$nn.idx[, -1]
  lab <- as.integer(factor(labels))
  mean(sapply(seq_len(nrow(nn)), function(i) mean(lab[nn[i, ]] == lab[i])))
}
```

## Standard UMAP (baseline)

```{r standard}
set.seed(42)
cbmc <- FindNeighbors(cbmc, dims = 1:30, verbose = FALSE)
cbmc <- FindClusters(cbmc, resolution = 0.8, verbose = FALSE)
cbmc <- RunUMAP(cbmc, dims = 1:30, reduction.name = "umap.std", verbose = FALSE)

ari_std <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_std <- knn_purity(Embeddings(cbmc, "umap.std"), cbmc$protein_annotations)
```

## Weighted UMAP variants

```{r weighted}
set.seed(42)

# stdev weighting
cbmc <- RunWeightedNeighbors(cbmc, dims = 1:30, weight.by = "stdev",
                             prefix = "wt.sd", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "wt.sd_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunWeightedUMAP(cbmc, dims = 1:30, weight.by = "stdev",
                        reduction.name = "umap.sd", verbose = FALSE)
ari_sd <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_sd <- knn_purity(Embeddings(cbmc, "umap.sd"), cbmc$protein_annotations)

# stdev + log.scale
cbmc <- RunWeightedNeighbors(cbmc, dims = 1:30, weight.by = "stdev",
                             log.scale = TRUE, prefix = "wt_sdlog",
                             verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "wt_sdlog_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunWeightedUMAP(cbmc, dims = 1:30, weight.by = "stdev",
                        log.scale = TRUE, reduction.name = "umap.sd.log",
                        verbose = FALSE)
ari_sd_log <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_sd_log <- knn_purity(Embeddings(cbmc, "umap.sd.log"), cbmc$protein_annotations)

# prop.var, weight.factor = 0.5
cbmc <- RunWeightedNeighbors(cbmc, dims = 1:30, weight.by = "prop.var",
                             weight.factor = 0.5, prefix = "wt.pv05",
                             verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "wt.pv05_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunWeightedUMAP(cbmc, dims = 1:30, weight.by = "prop.var",
                        weight.factor = 0.5, reduction.name = "umap.pv05",
                        verbose = FALSE)
ari_pv05 <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_pv05 <- knn_purity(Embeddings(cbmc, "umap.pv05"), cbmc$protein_annotations)

# prop.var, weight.factor = 1
cbmc <- RunWeightedNeighbors(cbmc, dims = 1:30, weight.by = "prop.var",
                             prefix = "wt.pv", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "wt.pv_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunWeightedUMAP(cbmc, dims = 1:30, weight.by = "prop.var",
                        reduction.name = "umap.pv", verbose = FALSE)
ari_pv <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_pv <- knn_purity(Embeddings(cbmc, "umap.pv"), cbmc$protein_annotations)

# prop.var + log.scale
cbmc <- RunWeightedNeighbors(cbmc, dims = 1:30, weight.by = "prop.var",
                             log.scale = TRUE, prefix = "wt_pvlog",
                             verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "wt_pvlog_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunWeightedUMAP(cbmc, dims = 1:30, weight.by = "prop.var",
                        log.scale = TRUE, reduction.name = "umap.pv.log",
                        verbose = FALSE)
ari_pv_log <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_pv_log <- knn_purity(Embeddings(cbmc, "umap.pv.log"), cbmc$protein_annotations)

# stdev + Marchenko-Pastur filtering: zero out noise PCs, weight signal PCs by stdev
cbmc <- RunWeightedNeighbors(cbmc, dims = 1:30, weight.by = "stdev",
                             mp.filter = TRUE, prefix = "wt.mp",
                             verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "wt.mp_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunWeightedUMAP(cbmc, dims = 1:30, weight.by = "stdev",
                        mp.filter = TRUE, reduction.name = "umap.mp",
                        verbose = FALSE)
ari_mp <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_mp <- knn_purity(Embeddings(cbmc, "umap.mp"), cbmc$protein_annotations)
```

## Local PCA UMAP

`RunLocalPCANeighbors()` builds KNN/SNN graphs from per-neighbourhood PCA
distances and stores them for consistent clustering and visualisation.
Using the same `k.param` and `local.weight.by` for both the graph and the UMAP
step ensures clustering and layout share identical distances.

We test all three `local.weight.by` schemes plus `mp.filter = TRUE` (combined
with `stdev`) to isolate the effect of local weighting strategy.

```{r local-pca}
set.seed(42)

# stdev (default)
cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30,
                             local.weight.by = "stdev",
                             prefix = "lp", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30,
                        local.weight.by = "stdev",
                        reduction.name = "umap.lp", verbose = FALSE)
ari_lp     <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp     <- knn_purity(Embeddings(cbmc, "umap.lp"), cbmc$protein_annotations)

# prop.var (aggressive local weighting)
cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30,
                             local.weight.by = "prop.var",
                             prefix = "lp.pv", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp.pv_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30,
                        local.weight.by = "prop.var",
                        reduction.name = "umap.lp.pv", verbose = FALSE)
ari_lp_pv  <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp_pv  <- knn_purity(Embeddings(cbmc, "umap.lp.pv"), cbmc$protein_annotations)

# none (unweighted local PCA, isotropic)
cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30,
                             local.weight.by = "none",
                             prefix = "lp.none", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp.none_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30,
                        local.weight.by = "none",
                        reduction.name = "umap.lp.none", verbose = FALSE)
ari_lp_none <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp_none <- knn_purity(Embeddings(cbmc, "umap.lp.none"), cbmc$protein_annotations)

# stdev + mp.filter (noise-PC zeroing post-weighting)
cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30,
                             local.weight.by = "stdev", mp.filter = TRUE,
                             prefix = "lp.mp", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp.mp_snn", resolution = 0.8,
                     verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30,
                        local.weight.by = "stdev", mp.filter = TRUE,
                        reduction.name = "umap.lp.mp", verbose = FALSE)
ari_lp_mp  <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp_mp  <- knn_purity(Embeddings(cbmc, "umap.lp.mp"), cbmc$protein_annotations)
```

## Local PCA: parameter sweep

The `k.param` and `local.dims` parameters control two orthogonal aspects of
the local PCA metric:

- **`k.param`**: neighbourhood size.  Smaller values capture finer local
  structure but produce noisier SVDs; larger values smooth out local variation
  and converge toward a global metric.
- **`local.dims`**: how many local PC directions to retain.  Truncating to
  fewer directions discards the weakest local axes, which is an alternative
  (non-probabilistic) form of noise filtering.

We fix `local.weight.by = "stdev"` throughout and vary each parameter in turn.

```{r local-pca-sweep}
set.seed(42)

# ── k.param sweep (local.dims = k.param, stdev weighting) ──────────────────
cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 10L,
                             local.weight.by = "stdev",
                             prefix = "lp.k10", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp.k10_snn",
                     resolution = 0.8, verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 10L,
                        local.weight.by = "stdev",
                        reduction.name = "umap.lp.k10", verbose = FALSE)
ari_lp_k10 <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp_k10 <- knn_purity(Embeddings(cbmc, "umap.lp.k10"), cbmc$protein_annotations)

cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 20L,
                             local.weight.by = "stdev",
                             prefix = "lp.k20", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp.k20_snn",
                     resolution = 0.8, verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 20L,
                        local.weight.by = "stdev",
                        reduction.name = "umap.lp.k20", verbose = FALSE)
ari_lp_k20 <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp_k20 <- knn_purity(Embeddings(cbmc, "umap.lp.k20"), cbmc$protein_annotations)

cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 50L,
                             local.weight.by = "stdev",
                             prefix = "lp.k50", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp.k50_snn",
                     resolution = 0.8, verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 50L,
                        local.weight.by = "stdev",
                        reduction.name = "umap.lp.k50", verbose = FALSE)
ari_lp_k50 <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp_k50 <- knn_purity(Embeddings(cbmc, "umap.lp.k50"), cbmc$protein_annotations)

# ── local.dims sweep (k.param = 30, stdev weighting) ───────────────────────
cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30L,
                             local.dims = 5L, local.weight.by = "stdev",
                             prefix = "lp.ld5", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp.ld5_snn",
                     resolution = 0.8, verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30L,
                        local.dims = 5L, local.weight.by = "stdev",
                        reduction.name = "umap.lp.ld5", verbose = FALSE)
ari_lp_ld5 <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp_ld5 <- knn_purity(Embeddings(cbmc, "umap.lp.ld5"), cbmc$protein_annotations)

cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30L,
                             local.dims = 10L, local.weight.by = "stdev",
                             prefix = "lp.ld10", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp.ld10_snn",
                     resolution = 0.8, verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30L,
                        local.dims = 10L, local.weight.by = "stdev",
                        reduction.name = "umap.lp.ld10", verbose = FALSE)
ari_lp_ld10 <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp_ld10 <- knn_purity(Embeddings(cbmc, "umap.lp.ld10"), cbmc$protein_annotations)

cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30L,
                             local.dims = 15L, local.weight.by = "stdev",
                             prefix = "lp.ld15", verbose = FALSE)
cbmc <- FindClusters(cbmc, graph.name = "lp.ld15_snn",
                     resolution = 0.8, verbose = FALSE)
cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30L,
                        local.dims = 15L, local.weight.by = "stdev",
                        reduction.name = "umap.lp.ld15", verbose = FALSE)
ari_lp_ld15 <- adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations)
pur_lp_ld15 <- knn_purity(Embeddings(cbmc, "umap.lp.ld15"), cbmc$protein_annotations)
```

```{r local-pca-sweep-table}
sweep_results <- data.frame(
  Setting = c(
    "k=10  (local.dims=10)",
    "k=20  (local.dims=20)",
    "k=30  (local.dims=30, baseline)",
    "k=50  (local.dims=50)",
    "k=30, local.dims=5",
    "k=30, local.dims=10",
    "k=30, local.dims=15",
    "k=30, local.dims=30 (baseline)"
  ),
  Varied     = c(rep("k.param", 4), rep("local.dims", 4)),
  ARI        = c(ari_lp_k10, ari_lp_k20, ari_lp, ari_lp_k50,
                 ari_lp_ld5, ari_lp_ld10, ari_lp_ld15, ari_lp),
  kNN_purity = c(pur_lp_k10, pur_lp_k20, pur_lp, pur_lp_k50,
                 pur_lp_ld5, pur_lp_ld10, pur_lp_ld15, pur_lp)
)
knitr::kable(sweep_results, digits = 3,
  caption = "Local PCA parameter sweep (local.weight.by = 'stdev' throughout).")
```

```{r local-pca-sweep-plot, fig.height=7, fig.cap="Local PCA parameter sweep. Dashed line = standard UMAP ARI baseline."}
sweep_long <- tidyr::pivot_longer(sweep_results, cols = c("ARI", "kNN_purity"),
                                  names_to = "Metric", values_to = "Score")
sweep_long$Setting <- factor(sweep_long$Setting, levels = sweep_results$Setting)
sweep_long$Varied  <- factor(sweep_long$Varied,  levels = c("k.param", "local.dims"))

ggplot(sweep_long, aes(x = Setting, y = Score, fill = Varied)) +
  geom_col(show.legend = TRUE) +
  geom_hline(data = data.frame(
    Metric = c("ARI", "kNN_purity"), Score = c(ari_std, pur_std)),
    aes(yintercept = Score), linetype = "dashed", colour = "grey40") +
  facet_wrap(~Metric, scales = "free_y") +
  scale_fill_manual(values = c("k.param" = "#4393C3", "local.dims" = "#D6604D")) +
  coord_flip() +
  labs(x = NULL, y = NULL, fill = "Parameter varied",
       title = "Local PCA parameter sweep (stdev weighting)",
       subtitle = "Dashed line = standard UMAP baseline") +
  theme_bw(base_size = 12)
```

```{r local-pca-sweep-umaps, fig.height=14, fig.cap="k.param sweep (top, stdev weighting) and local.dims sweep (bottom, k=30, stdev weighting)."}
Idents(cbmc) <- "protein_annotations"
sweep_plots <- list(
  DimPlot(cbmc, reduction = "umap.lp.k10",  label = TRUE, repel = TRUE) +
    ggtitle(sprintf("k=10  ARI=%.3f",  ari_lp_k10))  + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp.k20",  label = TRUE, repel = TRUE) +
    ggtitle(sprintf("k=20  ARI=%.3f",  ari_lp_k20))  + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp",      label = TRUE, repel = TRUE) +
    ggtitle(sprintf("k=30 (baseline)  ARI=%.3f", ari_lp)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp.k50",  label = TRUE, repel = TRUE) +
    ggtitle(sprintf("k=50  ARI=%.3f",  ari_lp_k50))  + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp.ld5",  label = TRUE, repel = TRUE) +
    ggtitle(sprintf("local.dims=5  ARI=%.3f",  ari_lp_ld5))  + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp.ld10", label = TRUE, repel = TRUE) +
    ggtitle(sprintf("local.dims=10  ARI=%.3f", ari_lp_ld10)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp.ld15", label = TRUE, repel = TRUE) +
    ggtitle(sprintf("local.dims=15  ARI=%.3f", ari_lp_ld15)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp",      label = TRUE, repel = TRUE) +
    ggtitle(sprintf("local.dims=30 (baseline)  ARI=%.3f", ari_lp)) + NoLegend()
)
wrap_plots(sweep_plots, ncol = 2)
```

## Local PCA: weighting × parameter interaction

The previous sweeps held `local.weight.by = "stdev"` fixed.  Here we run
all three weighting schemes across the same k and `local.dims` grids to check
whether the interaction matters — i.e. whether a different weighting scheme
performs better at certain neighbourhood sizes or truncation levels.

```{r local-pca-wt-sweep}
set.seed(42)

# ── prop.var at k = 10, 20, 50 (k=30 already in lp.pv) ─────────────────────
for (cfg in list(
  list(k=10L, pfx="lp.k10.pv",  rname="umap.lp.k10.pv"),
  list(k=20L, pfx="lp.k20.pv",  rname="umap.lp.k20.pv"),
  list(k=50L, pfx="lp.k50.pv",  rname="umap.lp.k50.pv")
)) {
  cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = cfg$k,
                               local.weight.by = "prop.var",
                               prefix = cfg$pfx, verbose = FALSE)
  cbmc <- FindClusters(cbmc, graph.name = paste0(cfg$pfx, "_snn"),
                       resolution = 0.8, verbose = FALSE)
  cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = cfg$k,
                          local.weight.by = "prop.var",
                          reduction.name = cfg$rname, verbose = FALSE)
  assign(paste0("ari_lp_k", cfg$k, "_pv"),
         adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations))
  assign(paste0("pur_lp_k", cfg$k, "_pv"),
         knn_purity(Embeddings(cbmc, cfg$rname), cbmc$protein_annotations))
}

# ── none at k = 10, 20, 50 (k=30 already in lp.none) ───────────────────────
for (cfg in list(
  list(k=10L, pfx="lp.k10.none",  rname="umap.lp.k10.none"),
  list(k=20L, pfx="lp.k20.none",  rname="umap.lp.k20.none"),
  list(k=50L, pfx="lp.k50.none",  rname="umap.lp.k50.none")
)) {
  cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = cfg$k,
                               local.weight.by = "none",
                               prefix = cfg$pfx, verbose = FALSE)
  cbmc <- FindClusters(cbmc, graph.name = paste0(cfg$pfx, "_snn"),
                       resolution = 0.8, verbose = FALSE)
  cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = cfg$k,
                          local.weight.by = "none",
                          reduction.name = cfg$rname, verbose = FALSE)
  assign(paste0("ari_lp_k", cfg$k, "_none"),
         adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations))
  assign(paste0("pur_lp_k", cfg$k, "_none"),
         knn_purity(Embeddings(cbmc, cfg$rname), cbmc$protein_annotations))
}

# ── prop.var at local.dims = 5, 10, 15 (ld=30 already in lp.pv) ─────────────
for (cfg in list(
  list(ld=5L,  pfx="lp.ld5.pv",  rname="umap.lp.ld5.pv"),
  list(ld=10L, pfx="lp.ld10.pv", rname="umap.lp.ld10.pv"),
  list(ld=15L, pfx="lp.ld15.pv", rname="umap.lp.ld15.pv")
)) {
  cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30L,
                               local.dims = cfg$ld, local.weight.by = "prop.var",
                               prefix = cfg$pfx, verbose = FALSE)
  cbmc <- FindClusters(cbmc, graph.name = paste0(cfg$pfx, "_snn"),
                       resolution = 0.8, verbose = FALSE)
  cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30L,
                          local.dims = cfg$ld, local.weight.by = "prop.var",
                          reduction.name = cfg$rname, verbose = FALSE)
  assign(paste0("ari_lp_ld", cfg$ld, "_pv"),
         adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations))
  assign(paste0("pur_lp_ld", cfg$ld, "_pv"),
         knn_purity(Embeddings(cbmc, cfg$rname), cbmc$protein_annotations))
}

# ── none at local.dims = 5, 10, 15 (ld=30 already in lp.none) ───────────────
for (cfg in list(
  list(ld=5L,  pfx="lp.ld5.none",  rname="umap.lp.ld5.none"),
  list(ld=10L, pfx="lp.ld10.none", rname="umap.lp.ld10.none"),
  list(ld=15L, pfx="lp.ld15.none", rname="umap.lp.ld15.none")
)) {
  cbmc <- RunLocalPCANeighbors(cbmc, dims = 1:30, k.param = 30L,
                               local.dims = cfg$ld, local.weight.by = "none",
                               prefix = cfg$pfx, verbose = FALSE)
  cbmc <- FindClusters(cbmc, graph.name = paste0(cfg$pfx, "_snn"),
                       resolution = 0.8, verbose = FALSE)
  cbmc <- RunLocalPCAUMAP(cbmc, dims = 1:30, k.param = 30L,
                          local.dims = cfg$ld, local.weight.by = "none",
                          reduction.name = cfg$rname, verbose = FALSE)
  assign(paste0("ari_lp_ld", cfg$ld, "_none"),
         adjustedRandIndex(cbmc$seurat_clusters, cbmc$protein_annotations))
  assign(paste0("pur_lp_ld", cfg$ld, "_none"),
         knn_purity(Embeddings(cbmc, cfg$rname), cbmc$protein_annotations))
}
```

```{r local-pca-wt-heatmap, fig.height=6, fig.cap="ARI heatmaps: weighting scheme × k.param (left) and weighting scheme × local.dims (right, k=30). Labels show ARI; higher is better."}
# ── Build grid data frames ──────────────────────────────────────────────────
wt_k_grid <- expand.grid(
  k      = c(10, 20, 30, 50),
  scheme = c("stdev", "prop.var", "none"),
  stringsAsFactors = FALSE
)
wt_k_grid$ARI <- c(
  ari_lp_k10,    ari_lp_k20,    ari_lp,    ari_lp_k50,
  ari_lp_k10_pv, ari_lp_k20_pv, ari_lp_pv, ari_lp_k50_pv,
  ari_lp_k10_none, ari_lp_k20_none, ari_lp_none, ari_lp_k50_none
)

wt_ld_grid <- expand.grid(
  local.dims = c(5, 10, 15, 30),
  scheme     = c("stdev", "prop.var", "none"),
  stringsAsFactors = FALSE
)
wt_ld_grid$ARI <- c(
  ari_lp_ld5,    ari_lp_ld10,    ari_lp_ld15,    ari_lp,
  ari_lp_ld5_pv, ari_lp_ld10_pv, ari_lp_ld15_pv, ari_lp_pv,
  ari_lp_ld5_none, ari_lp_ld10_none, ari_lp_ld15_none, ari_lp_none
)

# ── Plot ────────────────────────────────────────────────────────────────────
ph_k <- ggplot(wt_k_grid, aes(x = scheme, y = factor(k), fill = ARI)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.3f", ARI)), size = 3.5) +
  scale_fill_distiller(palette = "RdYlGn", direction = 1,
                       limits = range(c(wt_k_grid$ARI, wt_ld_grid$ARI))) +
  labs(x = "local.weight.by", y = "k.param",
       title = "k.param × weighting") +
  theme_bw(base_size = 12)

ph_ld <- ggplot(wt_ld_grid, aes(x = scheme, y = factor(local.dims), fill = ARI)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.3f", ARI)), size = 3.5) +
  scale_fill_distiller(palette = "RdYlGn", direction = 1,
                       limits = range(c(wt_k_grid$ARI, wt_ld_grid$ARI))) +
  labs(x = "local.weight.by", y = "local.dims",
       title = "local.dims × weighting  (k=30)") +
  theme_bw(base_size = 12)

ph_k | ph_ld
```

```{r wt-interaction-summary}
best_combo_k  <- wt_k_grid[which.max(wt_k_grid$ARI), ]
best_combo_ld <- wt_ld_grid[which.max(wt_ld_grid$ARI), ]
```

Across the full k × weighting grid, the highest ARI is
`r round(best_combo_k$ARI, 3)` at `k = `r best_combo_k$k``,
`local.weight.by = "`r best_combo_k$scheme`"`.

Across the local.dims × weighting grid, the highest ARI is
`r round(best_combo_ld$ARI, 3)` at `local.dims = `r best_combo_ld$local.dims``,
`local.weight.by = "`r best_combo_ld$scheme`"`.

If the best weighting scheme is consistent across all k values (each column in
the heatmap ranks the same way), the scheme choice dominates.  If the best
scheme changes with k, there is an interaction and neither parameter can be
tuned in isolation.

## Results summary

```{r results-table}
results <- data.frame(
  Method = c(
    "standard (none)",
    "stdev",
    "stdev + log.scale",
    "prop.var  wf=0.5",
    "prop.var  wf=1",
    "prop.var + log.scale",
    "stdev + mp.filter",
    "local PCA (stdev)",
    "local PCA (prop.var)",
    "local PCA (none)",
    "local PCA (stdev + mp.filter)"
  ),
  ARI       = c(ari_std, ari_sd, ari_sd_log, ari_pv05, ari_pv, ari_pv_log,
                ari_mp, ari_lp, ari_lp_pv, ari_lp_none, ari_lp_mp),
  kNN_purity = c(pur_std, pur_sd, pur_sd_log, pur_pv05, pur_pv, pur_pv_log,
                 pur_mp, pur_lp, pur_lp_pv, pur_lp_none, pur_lp_mp)
)
knitr::kable(results, digits = 3,
  caption = "Clustering quality vs protein labels on cbmc CITE-seq (n ≈ 7,699 cells).")
```

```{r results-plot, fig.height=8, fig.cap="ARI and k-NN purity for each method. Dashed line = standard UMAP baseline."}
library(tidyr)

res_long <- tidyr::pivot_longer(results, cols = c("ARI", "kNN_purity"),
                                names_to = "Metric", values_to = "Score")
res_long$Method <- factor(res_long$Method, levels = results$Method)

baseline <- data.frame(
  Metric = c("ARI", "kNN_purity"),
  Score  = c(ari_std, pur_std)
)

ggplot(res_long[!is.na(res_long$Score), ],
       aes(x = Method, y = Score, fill = Method)) +
  geom_col(show.legend = FALSE) +
  geom_hline(data = baseline, aes(yintercept = Score),
             linetype = "dashed", colour = "grey40") +
  facet_wrap(~Metric, scales = "free_y") +
  scale_fill_viridis_d(option = "C") +
  coord_flip() +
  labs(x = NULL, y = NULL,
       title = "wUMAP benchmark on cbmc CITE-seq",
       subtitle = "Dashed line = standard UMAP baseline") +
  theme_bw(base_size = 12)
```

## UMAP layouts coloured by protein label

```{r umap-grid, fig.height=24, fig.cap="All methods coloured by ADT-derived protein label."}
Idents(cbmc) <- "protein_annotations"

plots <- list(
  DimPlot(cbmc, reduction = "umap.std",   label = TRUE, repel = TRUE) +
    ggtitle(sprintf("standard  ARI=%.3f", ari_std))    + NoLegend(),
  DimPlot(cbmc, reduction = "umap.sd",    label = TRUE, repel = TRUE) +
    ggtitle(sprintf("stdev  ARI=%.3f", ari_sd))        + NoLegend(),
  DimPlot(cbmc, reduction = "umap.sd.log", label = TRUE, repel = TRUE) +
    ggtitle(sprintf("stdev+log  ARI=%.3f", ari_sd_log))+ NoLegend(),
  DimPlot(cbmc, reduction = "umap.pv05",  label = TRUE, repel = TRUE) +
    ggtitle(sprintf("prop.var wf=0.5  ARI=%.3f", ari_pv05)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.pv",    label = TRUE, repel = TRUE) +
    ggtitle(sprintf("prop.var wf=1  ARI=%.3f", ari_pv)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.pv.log", label = TRUE, repel = TRUE) +
    ggtitle(sprintf("prop.var+log  ARI=%.3f", ari_pv_log)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.mp",    label = TRUE, repel = TRUE) +
    ggtitle(sprintf("stdev+mp.filter  ARI=%.3f", ari_mp)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp",    label = TRUE, repel = TRUE) +
    ggtitle(sprintf("local PCA stdev  ARI=%.3f", ari_lp)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp.pv", label = TRUE, repel = TRUE) +
    ggtitle(sprintf("local PCA prop.var  ARI=%.3f", ari_lp_pv)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp.none", label = TRUE, repel = TRUE) +
    ggtitle(sprintf("local PCA none  ARI=%.3f", ari_lp_none)) + NoLegend(),
  DimPlot(cbmc, reduction = "umap.lp.mp", label = TRUE, repel = TRUE) +
    ggtitle(sprintf("local PCA stdev+mp.filter  ARI=%.3f", ari_lp_mp)) + NoLegend()
)

wrap_plots(plots, ncol = 2)
```

## Interpretation

On this PBMC-like dataset, aggressive `prop.var` weighting (`wf=1`) **reduces**
clustering quality (ARI 0.548 → 0.318). PC 1 in scRNA-seq captures
the monocyte–lymphocyte axis so strongly that further emphasising it collapses
finer lymphocyte subtypes together.

Key takeaways:

- **`stdev`** is the safest global default — only a marginal ARI change, same
  k-NN purity as standard.
- **`prop.var` with `weight.factor = 0.5`** (partial blend) avoids the worst
  of over-emphasising PC 1.
- **`log.scale = TRUE`** partially mitigates PC 1 dominance and recovers some
  ARI compared to full `prop.var`.
- **`mp.filter = TRUE`** (global: combined with `stdev`) automatically zeros
  out noise PCs whose variance is below the Marchenko–Pastur ceiling. On
  datasets where most PCs are genuine signal it behaves like the base scheme;
  on noisier datasets it can sharpen clusters.

### Local PCA parameter sweep

```{r sweep-summary}
k_aris  <- c(ari_lp_k10, ari_lp_k20, ari_lp, ari_lp_k50)
k_vals  <- c(10, 20, 30, 50)
best_k  <- k_vals[which.max(k_aris)]

ld_aris <- c(ari_lp_ld5, ari_lp_ld10, ari_lp_ld15, ari_lp)
ld_vals <- c(5, 10, 15, 30)
best_ld <- ld_vals[which.max(ld_aris)]
```

For the **k.param sweep**, `k = `r best_k`` achieves the highest ARI
(`r round(k_aris[k_vals == best_k], 3)`).

- If ARI improves at smaller k, finer local structure (e.g. subtle sub-types)
  is being captured.
- If ARI is flat across k, the local SVD is not providing information beyond
  a global metric on this dataset.
- If ARI degrades at very small k (e.g. k=10), the neighbourhood is too small
  for a stable SVD and the metric becomes noisy.

For the **local.dims sweep** (k=30), `local.dims = `r best_ld`` achieves the
highest ARI (`r round(ld_aris[ld_vals == best_ld], 3)`).

- If fewer local dims performs better, the weaker local directions add noise
  rather than signal — truncation acts as implicit regularisation.
- If all dims perform similarly, the choice of `local.dims` is not a sensitive
  parameter for this dataset.


### Local PCA weighting schemes (at k=30)

The following table and summary cover the four weighting variants at the
default k=30, comparing stdev, prop.var, none, and stdev+mp.filter.

| Setting | What it does |
|---|---|
| `local.weight.by = "stdev"` | Weights each local PC by its local std dev — gently emphasises dominant local directions |
| `local.weight.by = "prop.var"` | Weights by proportion of local variance — more aggressive emphasis on the leading local direction |
| `local.weight.by = "none"` | Equal weight to all local PCs — isotropic local PCA distance |
| `stdev + mp.filter = TRUE` | Runs stdev weighting then zeros local PCs below the per-neighbourhood MP noise threshold |

```{r local-pca-summary}
lp_results <- results[results$Method %in% c(
  "local PCA (stdev)", "local PCA (prop.var)",
  "local PCA (none)", "local PCA (stdev + mp.filter)"
), ]
best_lp  <- lp_results$Method[which.max(lp_results$ARI)]
worst_lp <- lp_results$Method[which.min(lp_results$ARI)]
```

Among the local PCA variants, **`r best_lp`** achieves the highest ARI
(`r round(lp_results$ARI[lp_results$Method == best_lp], 3)`) and
**`r worst_lp`** the lowest
(`r round(lp_results$ARI[lp_results$Method == worst_lp], 3)`).

In general:

- If all local PCA variants score similarly to each other, the local graph
  structure (rather than the weighting scheme) is the dominant factor; the
  `stdev` default is sufficient.
- If `prop.var` outperforms `stdev`, the dataset has strong anisotropic local
  structure and aggressive local direction emphasis pays off.
- If `none` is competitive, the data has broadly isotropic neighbourhoods and
  local weighting adds little.
- If `stdev + mp.filter` outperforms plain `stdev`, per-neighbourhood noise PCs
  were degrading the distance estimate and zeroing them helps.

Comparing local PCA methods against global methods: local PCA will typically
match or surpass global weighting on trajectory-like or heterogeneous datasets,
but on PBMC data with a dominant monocyte–lymphocyte axis the global and local
distances may perform similarly.

**Note on interpretation**: all wUMAP weighting schemes are heuristic
modifications of the input distance metric — they are not derived from a
statistical model of scRNA-seq noise.  On this PBMC dataset the differences
between methods tend to be small.  The benchmark measures whether the
heuristics help; results will vary by dataset.

**Recommendation**: the default `weight.by = "stdev"` / `local.weight.by =
"stdev"` applies the mildest transformation and is a reasonable starting
point.  Use `mp.filter = TRUE` to additionally zero PCs below the
Marchenko-Pastur noise ceiling (a principled threshold, though calibrated for
i.i.d. Gaussian data and therefore approximate on scRNA-seq).  Try
`RunLocalPCANeighbors()` + `RunLocalPCAUMAP()` when you suspect the data has
locally anisotropic structure (trajectories, gradients), but verify empirically
that it helps.

## Trajectory dataset: C. elegans embryo

To test the claim that wUMAP performs better on trajectory-like or highly heterogeneous datasets, we benchmark on the `celegans.embryo` dataset from `SeuratData`, which contains a developmental trajectory across worm embryo stages.

```{r embryo-setup, message=TRUE, warning=TRUE}
# Install and load celegans.embryo dataset from SeuratData if not already present
if (!"celegans.embryo" %in% SeuratData::AvailableData()$Dataset) {
  SeuratData::InstallData("celegans.embryo", ask = FALSE)
}
library(SeuratData)
data(celegans.embryo)
embryo <- UpdateSeuratObject(celegans.embryo)
embryo <- NormalizeData(embryo)
embryo <- FindVariableFeatures(embryo, nfeatures = 2000)
embryo <- ScaleData(embryo)
embryo <- RunPCA(embryo, npcs = 30)
# Use published cell type labels for benchmarking
Idents(embryo) <- embryo$cell.type
```

### wUMAP baseline: weight.none

```{r embryo-baseline, warning=TRUE, message=TRUE}
embryo <- RunWeightedNeighbors(embryo, dims = 1:30, weight.by = "none",
                               reduction.name = "wt.none.pca",
                               prefix = "wt.none")
embryo <- FindClusters(embryo, graph.name = "wt.none_snn",
                       cluster.name = "wt.none_clusters",
                       resolution = 0.8, verbose = FALSE)
embryo <- RunWeightedUMAP(embryo, graph = "wt.none_nn",
                          reduction.name = "umap.none",
                          seed.use = 42)
```

### wUMAP (stdev weighting)

```{r embryo-wumap, warning=TRUE, message=TRUE}
embryo <- RunWeightedNeighbors(embryo, dims = 1:30, weight.by = "stdev",
                               reduction.name = "wt.stdev.pca",
                               prefix = "wt.stdev")
embryo <- FindClusters(embryo, graph.name = "wt.stdev_snn",
                       cluster.name = "wt.stdev_clusters",
                       resolution = 0.8, verbose = FALSE)
embryo <- RunWeightedUMAP(embryo, graph = "wt.stdev_nn",
                          reduction.name = "umap.stdev",
                          seed.use = 42)
```

### Local PCA UMAP

```{r embryo-lp, warning=TRUE, message=TRUE}
embryo <- RunLocalPCANeighbors(embryo, dims = 1:30, k.param = 30,
                               local.weight.by = "none",
                               prefix = "lp.none")
embryo <- FindClusters(embryo, graph.name = "lp.none_snn",
                       cluster.name = "lp.none_clusters",
                       resolution = 0.8, verbose = FALSE)
embryo <- RunLocalPCAUMAP(embryo, dims = 1:30, k.param = 30,
                          local.weight.by = "none",
                          reduction.name = "umap.lp.none",
                          seed.use = 42)

embryo <- RunLocalPCANeighbors(embryo, dims = 1:30, k.param = 30,
                               local.weight.by = "stdev",
                               prefix = "lp.stdev")
embryo <- FindClusters(embryo, graph.name = "lp.stdev_snn",
                       cluster.name = "lp.stdev_clusters",
                       resolution = 0.8, verbose = FALSE)
embryo <- RunLocalPCAUMAP(embryo, dims = 1:30, k.param = 30,
                          local.weight.by = "stdev",
                          reduction.name = "umap.lp.stdev",
                          seed.use = 42)
```

### Visualization

```{r embryo-umap-plot, fig.height=10}
p1 <- DimPlot(embryo, reduction = "umap.none", group.by = "wt.none_clusters", label = TRUE) +
  ggtitle("wUMAP weight.none") + NoLegend()
p2 <- DimPlot(embryo, reduction = "umap.stdev", group.by = "wt.stdev_clusters", label = TRUE) +
  ggtitle("wUMAP stdev") + NoLegend()
p3 <- DimPlot(embryo, reduction = "umap.lp.none", group.by = "lp.none_clusters", label = TRUE) +
  ggtitle("Local PCA none") + NoLegend()
p4 <- DimPlot(embryo, reduction = "umap.lp.stdev", group.by = "lp.stdev_clusters", label = TRUE) +
  ggtitle("Local PCA stdev") + NoLegend()
wrap_plots(p1, p2, p3, p4, ncol = 2)
```

### Differentiation stage on embryo UMAPs

```{r embryo-differentiation, fig.height=10}
# Colour by developmental time bins and by continuous embryo time
p1 <- DimPlot(embryo, reduction = "umap.none", group.by = "embryo.time.bin",
              label = FALSE) + ggtitle("wUMAP weight.none") + NoLegend()
p2 <- DimPlot(embryo, reduction = "umap.stdev", group.by = "embryo.time.bin",
              label = FALSE) + ggtitle("wUMAP stdev") + NoLegend()
p3 <- DimPlot(embryo, reduction = "umap.lp.none", group.by = "embryo.time.bin",
              label = FALSE) + ggtitle("Local PCA none") + NoLegend()
p4 <- DimPlot(embryo, reduction = "umap.lp.stdev", group.by = "embryo.time.bin",
              label = FALSE) + ggtitle("Local PCA stdev") + NoLegend()
wrap_plots(p1, p2, p3, p4, ncol = 2)
```

```{r embryo-differentiation-continuous, fig.height=10}
p1 <- FeaturePlot(embryo, features = "embryo.time", reduction = "umap.none",
                  pt.size = 0.5) + ggtitle("wUMAP weight.none")
p2 <- FeaturePlot(embryo, features = "embryo.time", reduction = "umap.stdev",
                  pt.size = 0.5) + ggtitle("wUMAP stdev")
p3 <- FeaturePlot(embryo, features = "embryo.time", reduction = "umap.lp.none",
                  pt.size = 0.5) + ggtitle("Local PCA none")
p4 <- FeaturePlot(embryo, features = "embryo.time", reduction = "umap.lp.stdev",
                  pt.size = 0.5) + ggtitle("Local PCA stdev")
wrap_plots(p1, p2, p3, p4, ncol = 2)
```

### Quantitative comparison

```{r embryo-ari}
library(mclust)
ari_none   <- adjustedRandIndex(embryo$cell.type, embryo$wt.none_clusters)
ari_stdev  <- adjustedRandIndex(embryo$cell.type, embryo$wt.stdev_clusters)
ari_lpnone <- adjustedRandIndex(embryo$cell.type, embryo$lp.none_clusters)
ari_lpstdv <- adjustedRandIndex(embryo$cell.type, embryo$lp.stdev_clusters)
data.frame(
  Method = c("wUMAP weight.none", "wUMAP stdev", "Local PCA none", "Local PCA stdev"),
  ARI = c(ari_none, ari_stdev, ari_lpnone, ari_lpstdv)
)
```

### Interpretation

On this trajectory dataset, compare the ARI and visual separation of cell types between standard UMAP and wUMAP. If wUMAP better preserves the developmental trajectory or improves ARI, it supports the claim. Otherwise, the benefit may be context-dependent.
