---
title: "Variance-Weighted UMAP with wUMAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variance-Weighted UMAP with wUMAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 10,
  fig.height = 5,
  dpi       = 150,
  message   = FALSE,
  warning   = FALSE
)
```

## Overview

Standard UMAP treats every principal component (PC) equally. In scRNA-seq,
early PCs capture far more biological variation than later ones — yet a
standard UMAP over PCs 1–50 weights PC 50 the same as PC 1.

**wUMAP** scales each PC axis by a weight derived from its variance contribution
before handing the embedding to UMAP, so biologically informative PCs dominate
cell distances in the final layout.

## Setup

```{r libs}
library(Seurat)
library(SeuratData)
library(ggplot2)
library(wUMAP)
library(patchwork)
```

## Data

We use the classic PBMC 3k dataset from `SeuratData`, which includes PCA,
cell-type annotations, and standard clustering already applied.

```{r data}
data(pbmc3k.final)
pbmc <- UpdateSeuratObject(pbmc3k.final)
Idents(pbmc) <- "seurat_annotations"
```

## Standard vs weighted UMAP

`RunWeightedUMAP()` is a drop-in replacement for `RunUMAP()`. Setting
`weight.by = "none"` reproduces the standard result; `weight.by = "prop.var"`
(the default) scales each PC by its proportion of explained variance.

```{r run-umap}
set.seed(42)
pbmc <- RunWeightedUMAP(pbmc, dims = 1:50, weight.by = "none",
                        reduction.name = "umap.std",     verbose = FALSE)

pbmc <- RunWeightedUMAP(pbmc, dims = 1:50, weight.by = "prop.var",
                        reduction.name = "umap.wt",      verbose = FALSE)
```

```{r plot-comparison, fig.cap="Left: standard UMAP. Right: variance-weighted UMAP."}
p1 <- DimPlot(pbmc, reduction = "umap.std", label = TRUE, repel = TRUE) +
  ggtitle("Standard UMAP") + NoLegend()
p2 <- DimPlot(pbmc, reduction = "umap.wt",  label = TRUE, repel = TRUE) +
  ggtitle("Weighted UMAP (prop.var)") + NoLegend()
p1 | p2
```

## Weighting schemes

Three schemes are available via `weight.by`:

| Value | Weight formula | Effect |
|---|---|---|
| `"prop.var"` | $w_i = \sigma_i^2 / \sum \sigma^2$ | Early PCs strongly dominate |
| `"stdev"` | $w_i = \sigma_i / \sum \sigma$ | Gentler emphasis on early PCs |
| `"mp"` | $w_i = \max(0,\,\sigma_i^2 - \lambda_{\max})$ normalised | Keeps only PCs above the Marchenko–Pastur noise floor |
| `"none"` | $w_i = 1/d$ | Standard UMAP |

The `"mp"` option applies the **Marchenko–Pastur** (MP) law from random matrix
theory to determine which PCs carry signal above the noise floor. Given $n$
cells and $p$ features, the theoretical noise ceiling is
$$\lambda_{\max} = \left(1 + \sqrt{\frac{p}{n}}\right)^2$$
Only PCs whose variance $\sigma_i^2 > \lambda_{\max}$ receive non-zero weight;
all others are treated as noise and zeroed out. If no PC clears the threshold
(e.g. on tiny datasets) the function falls back to `"prop.var"` with a warning.

```{r weight-by-stdev}
pbmc <- RunWeightedUMAP(pbmc, dims = 1:50, weight.by = "stdev",
                        reduction.name = "umap.wt.sd", verbose = FALSE)
```

```{r plot-stdev, fig.cap="Proportional-variance vs standard-deviation weighting."}
p3 <- DimPlot(pbmc, reduction = "umap.wt.sd", label = TRUE, repel = TRUE) +
  ggtitle("Weighted UMAP (stdev)") + NoLegend()
p2 | p3
```

## Marchenko–Pastur weighting

`weight.by = "mp"` uses random matrix theory to identify PCs that carry
genuine signal above the noise floor predicted for a random matrix of the
same dimensions. Only those PCs receive non-zero weight.

```{r weight-by-mp}
pbmc <- RunWeightedUMAP(pbmc, dims = 1:50, weight.by = "mp",
                        reduction.name = "umap.mp", verbose = FALSE)
```

We can inspect how many PCs were retained (non-zero weight):

```{r mp-weights}
mp_weights <- Misc(pbmc[["umap.mp"]])$weights
cat(sprintf("%d / %d PCs above noise floor (non-zero weight)\n",
            sum(mp_weights > 0), length(mp_weights)))
barplot(mp_weights, names.arg = seq_along(mp_weights),
        xlab = "PC", ylab = "MP weight",
        main = "Marchenko-Pastur weights (zero = noise)",
        col = ifelse(mp_weights > 0, "#4393C3", "#D1D1D1"),
        border = NA, las = 1)
```

```{r plot-mp, fig.cap="MP weighting vs proportional-variance weighting."}
p_mp <- DimPlot(pbmc, reduction = "umap.mp", label = TRUE, repel = TRUE) +
  ggtitle("Weighted UMAP (mp)") + NoLegend()
p2 | p_mp
```

## Blending with `weight.factor`

`weight.factor` (0–1) continuously blends between the unweighted (`0`) and
fully weighted (`1`) embedding:

```{r weight-factor}
pbmc <- RunWeightedUMAP(pbmc, dims = 1:50, weight.by = "prop.var",
                        weight.factor = 0.25, reduction.name = "umap.wf25",
                        verbose = FALSE)
pbmc <- RunWeightedUMAP(pbmc, dims = 1:50, weight.by = "prop.var",
                        weight.factor = 0.75, reduction.name = "umap.wf75",
                        verbose = FALSE)
```

```{r plot-blend, fig.cap="weight.factor controls the blend between standard and fully weighted."}
pa <- DimPlot(pbmc, reduction = "umap.std",  label = TRUE, repel = TRUE) +
  ggtitle("weight.factor = 0") + NoLegend()
pb <- DimPlot(pbmc, reduction = "umap.wf25", label = TRUE, repel = TRUE) +
  ggtitle("weight.factor = 0.25") + NoLegend()
pc <- DimPlot(pbmc, reduction = "umap.wf75", label = TRUE, repel = TRUE) +
  ggtitle("weight.factor = 0.75") + NoLegend()
pd <- DimPlot(pbmc, reduction = "umap.wt",   label = TRUE, repel = TRUE) +
  ggtitle("weight.factor = 1") + NoLegend()
(pa | pb) / (pc | pd)
```

## Log-scale weights

For datasets where PC 1 explains dramatically more variance than all others,
`log.scale = TRUE` first applies `log1p()` to the weights before normalising.
This compresses the dynamic range so intermediate PCs still contribute to the
layout.

```{r log-scale}
pbmc <- RunWeightedUMAP(pbmc, dims = 1:50, weight.by = "prop.var",
                        log.scale = TRUE, reduction.name = "umap.log",
                        verbose = FALSE)
```

```{r plot-log, fig.cap="Standard (left), fully weighted (centre), log-scaled weights (right)."}
DimPlot(pbmc, reduction = "umap.std", label = TRUE, repel = TRUE) + ggtitle("Standard") + NoLegend() |
DimPlot(pbmc, reduction = "umap.wt",  label = TRUE, repel = TRUE) + ggtitle("prop.var") + NoLegend() |
DimPlot(pbmc, reduction = "umap.log", label = TRUE, repel = TRUE) + ggtitle("prop.var + log.scale") + NoLegend()
```

## Consistent clustering and UMAP with `RunWeightedNeighbors()`

A common pitfall: clustering on one KNN graph while visualising on another.
`RunWeightedNeighbors()` stores the weighted embedding and builds both the KNN
and SNN graphs in one step, so clustering and UMAP share *exactly* the same
nearest-neighbour structure.

```{r weighted-neighbors}
pbmc <- RunWeightedNeighbors(pbmc, dims = 1:50, weight.by = "prop.var",
                             prefix = "wt", verbose = FALSE)

pbmc <- FindClusters(pbmc, graph.name = "wt_snn", resolution = 0.5,
                     verbose = FALSE)

pbmc <- RunWeightedUMAP(pbmc, graph = "wt_nn", reduction.name = "umap.consistent",
                        verbose = FALSE)
```

```{r plot-consistent, fig.cap="Clustering and UMAP both derived from the same weighted KNN graph."}
DimPlot(pbmc, reduction = "umap.consistent", label = TRUE, repel = TRUE) +
  ggtitle("Consistent weighted clustering + UMAP") + NoLegend()
```

## Local PCA UMAP

`RunLocalPCAUMAP()` is a more principled alternative to global variance
weighting. Rather than rescaling PC axes globally, it measures the distance
between every pair of neighbours in a **locally-fitted PCA basis** — capturing
the predominant direction of variation in each cell's neighbourhood (e.g. the
tangent of a trajectory) and de-emphasising transverse noise.

**Algorithm:**

1. Find the global $k$ nearest neighbours of every cell in PCA space (RANN).
2. For each cell $i$, centre its $k$-neighbourhood and compute a compact SVD.
3. Re-express displacement vectors to neighbours in the local `local.dims`
   principal directions.
4. Report the Euclidean norm in that local basis as the refined distance.
5. Pass the $n \times k$ index/distance matrix directly to `uwot::umap` as
   a precomputed k-NN graph.

```{r local-pca-umap}
set.seed(42)
pbmc <- RunLocalPCAUMAP(pbmc, dims = 1:30, k.param = 30,
                        reduction.name = "lp.umap", verbose = FALSE)
```

```{r plot-local-pca, fig.cap="Standard UMAP (left) vs local PCA UMAP (right)."}
p_std <- DimPlot(pbmc, reduction = "umap.std",  label = TRUE, repel = TRUE) +
  ggtitle("Standard UMAP") + NoLegend()
p_lp  <- DimPlot(pbmc, reduction = "lp.umap",   label = TRUE, repel = TRUE) +
  ggtitle("Local PCA UMAP") + NoLegend()
p_std | p_lp
```

The local PCA approach can reveal trajectory-like structure that is compressed
in standard UMAP because the distance metric is insensitive to the predominant
local direction of variation.

## Further reading

For a quantitative evaluation of all weighting schemes (including `"mp"` and
`RunLocalPCAUMAP`) against an independent protein-based ground truth (cbmc
CITE-seq), see the [Benchmark](benchmark.html) article.

